Generalized Tractability

### 2023-05-06 GeneralizedTractability_230501.tex
Updates from Overleaf

For this case where arbitrary linear functionals are allowed, the optimal solution is
\[
\APP_d(f,\varepsilon) = \sum_{i=1}^n \SOL(u_i) \ip[\cf_d]{f}{u_i},
\]
and the information complexity of our linear problem is
\begin{equation}\label{eq:comp_def}
\comp(\varepsilon, d) = \min \{n \in \natzero : \lambda_{n+1, d} \le \varepsilon\}, \qquad \comp:(0,\infty) \times \naturals \to \natzero.
\end{equation}

@@

Certain common notions of tractability are covered by the general situation described above.
{\small
\begin{equation*}
	\begin{array}{r@{\qquad}c@{\qquad}cc}
		\text{Tractability type} & T(\varepsilon^{-1},d,\vp)
		& T(0,d,\vp)\\
		\toprule
		\text{strong polynomial algebraic} & \max(1,\varepsilon^{-p}) & 1\\
		\text{polynomial algebraic} & \max(1,\varepsilon^{-p})d^{q} & d^{q}\\
		\text{strong polynomial exponential} &  [\max(1,\log(1 + \varepsilon^{-1})]^p & 1\\
		\text{polynomial exponential } &
		[\max(1,\log(1 + \varepsilon^{-1}))]^p  d^{q} & d^{q} \\
        \text{quasi-polynomial} & 
        \exp\{p(1+\log(\max(1,\varepsilon^{-1}))(1+\log(d))\} &
        \exp\{p(1+\log(d))\}
	\end{array}
\end{equation*}}
Here above, $\vp = (p,q)$ in some cases and $\vp$ is a scalar in other cases. A general reference
for results on algebraic tractability is \cite{NW08}, for exponential tractability we refer to \cite{HKW19} and \cite{KW19}.

@@

END OF THE ALTERNATIVE PROOF.]

@@

\begin{proof}
    \textbf{Sufficient Condition:}\\
Suppose that \eqref{eq:tractiff4} holds for some $\vp \ge 0$. By the argument in the proof of the sufficient condition for Theorem \ref{thm_main_strong_tract2}, it follows from \eqref{eq:compUP} with $d'=d$ that the information complexity has an upper bound of
\begin{equation*}
	 \comp(\varepsilon,d) \le  T(\varepsilon^{-1},d,\vp) \sum_{i=1}^\infty \frac{1}{T(\lambda_{i, d}^{-1},d,\vp)} + 1
\end{equation*}

@@

Let $J(\vp,d): = \lceil L_{\vp}\, T(0,d,\vp)\rceil$. By the argument in the proof of the sufficient condition for Theorem \ref{thm_main_strong_tract2}, it follows from \eqref{eq:compUP} with $d'=d$ that the information complexity has an upper bound of
\begin{equation*}
	 \comp(\varepsilon,d) \le  T(\varepsilon^{-1},d,\vp) \sum_{i=J(\vp,d)}^\infty \frac{1}{T(\lambda_{i, d}^{-1},d,\vp)} + J(\vp,d)
\end{equation*}
 Then this upper bound on the complexity can be further bounded above by
\begin{align*}
       \comp(\varepsilon,d)
       & \le T(\varepsilon^{-1},d,\vp) \Biggl [ 
       \underbrace{\sum_{i=J(\vp,d)}^\infty \frac{1}{T(\lambda_{i, d}^{-1},d,\vp)}}_{\le S_{\vp} \text{ by \eqref{eq:tractiff4}}}
       + \frac{J(\vp,d)}{T(\varepsilon^{-1},d, \vp)} \Biggr ] \\
       & \le T(\varepsilon^{-1},d,\vp)\left[ S_{\vp} + \frac{L_{\vp} T(0,d,\vp)+1}{T(\varepsilon^{-1},d, \vp)}\right]\\
       & \le T(\varepsilon^{-1},d,\vp)\left[S_{\vp} +\frac{L_{\vp} T(0,d,\vp)}{T(\varepsilon^{-1},d, \vp)}
       + \frac{1}{T(0,1, \vp)}\right]
       \qquad \text{by \eqref{eq:Tnondecreasing}}   \\
       & \le T(\varepsilon^{-1},d,\vp) 
       \underbrace{\left [  S_{\vp} + L_{\vp} + \frac{1}{T(0,1, \vp)}\right]}_{=:C_{\vp}}
       \qquad \text{by \eqref{eq:Tnondecreasing}}  
\end{align*}

@@

Let the tractability function, $T$,
be defined by
\[
 T(\varepsilon^{-1},d,p)= (\max\{\varepsilon^{-1},1\})^p
 \qquad \forall \varepsilon > 0, \,  d \in \naturals, \, p \ge 0.
\]
\fred{May we pull the exponent $p$ inside the $\max$?} Then Theorem \ref{thm_main_strong_tract2} yields the  equivalent condition for strong algebraic polynomial tractability:
\[
 \sup_{d\in\naturals} \sum_{i=L_p}^\infty (\min\{\lambda_{i,d},1\})^p < \infty \text{ for some } p \ge 0, L_p \in \naturals.
\]
Furthermore, the optimal $p$ is the infimum of all $p$ for which the above condition holds.
This essentially recovers the result in \cite[Theorem 5.1]{NW08}.
\end{example}

\begin{example}[Algebraic polynomial tractability]
Let the tractability function, $T$,
be defined by
\[
 T(\varepsilon^{-1},d,p,q)= (\max\{\varepsilon^{-1},1\})^p\, d^q
 \qquad \forall \varepsilon > 0, \,  d \in \naturals, \, p,q \ge 0.
\]
Then Theorem \ref{thm_main_tract2} yields the  equivalent condition for algebraic polynomial tractability:
\[
 \sup_{d\in\naturals} \sum_{i=L_{p,q} d^{q}}^\infty (\min\{\lambda_{i,d},1\})^p d^{-q} < \infty \text{ for some } p,q \ge 0, L_{p,q} \in \naturals.
\]
Furthermore, the set of optimal $(p,q)$ is a subset of
\begin{multline*}
\{(p^*,q^*) : \text{the above condition holds for all }(p,q) \in (p^*,\infty) \times (q^*,\infty) \\
\text{but not for any }(p,q) \in [0,p^*/2) \times [0,q^*/2)\}.
\end{multline*}
This essentially recovers the result on polynomial tractability in \cite[Theorem 5.1]{NW08} \fred{same theorem as above??, what about optimality in \cite[Theorem 5.1]{NW08}}.
\end{example}

\kachi{\begin{example}[EXP-strong polynomial tractability]
Let $ T :(0,\infty) \times \mathbb{N} \times [0,\infty)^s \rightarrow (0,\infty)$
be defined by
\[
 T(\varepsilon^{-1},d,\vp)= (1+\log\max\{\varepsilon^{-1},1\})^p.
\]
Then Theorem \ref{thm_main_tract2} yields the iff-condition
\[
 \sup_{d\in\naturals} \sum_{i=L_{\vp} d^{q_1}}^\infty (\min\{\lambda_{i,d},1\})^p  < \infty.
\]
\end{example}
\begin{example}[EXP-polynomial tractability]

Let $ T :(0,\infty) \times \mathbb{N} \times [0,\infty)^s \rightarrow (0,\infty)$
be defined by
\[
 T(\varepsilon^{-1},d,\vp)= (1+\log \max\{\varepsilon^{-1},1\})^p\, d^q.
\]
Then Theorem \ref{thm_main_tract2} yields the iff-condition
\[
 \sup_{d\in\naturals} \sum_{i=L_{\vp} d^{q_1}}^\infty (\min\{\lambda_{i,d},1\})^p d^{-q} < \infty.
\]
\end{example}}
\bigskip

%\begin{definition}[Quasi-polynomial Tractability]
 %   The problem is \emph{quasi-polynomially tractable}, if there exist two constants $C, t> 0$ such that
  %  \[
   % n(\varepsilon,d) \leq C\exp\{t(1+\ln(\varepsilon^{-1}))(1+\ln(d))\}
    %\] for all $\varepsilon \in (0,1)$ and $d\in\mathbb{N}$.
%\end{definition}

%\paragraph{Example 5: $(t,s)$-Weak tractability}
%Let $ T :(0,\infty) \times \mathbb{N} \times [0,\infty)^s \rightarrow (0,\infty)$. Suppose we have $T$-$(s,t)$- weak tractability, such that
%\[ \lim_{d+\varepsilon^{-1}\rightarrow\infty} \frac{\log \max(1,n(\varepsilon,d))}{d^t +(\max(1,\varepsilon^{-1}))^s} =0.\]
%Then Theorem \ref{thm:weak} yields the iff-condition
%\[
%\sup_{d\in\naturals} \sigma (c,d,s,\vp) \exp\left(-c\, T(0,d^t,\vp)\right)<\infty
%\]
%where $\displaystyle\sigma (c,d,s,\vp) =\sum_{i=1}^\infty \exp\left(-c\left(\lambda^{-1}_{i,d}\right)^{s/2}\right) $
\kachi{
\begin{example}[Quasi-$T$-polynomial tractability]
\begin{definition}
    As defined in \cite{GW11}, we say that a problem is \emph{ quasi-polynomially tractable (QPT)} if there exist two constants $C$, $t > 0$, such that
\begin{equation}\label{expqpt}
    n(\varepsilon,d) \leq C\exp\{t(1+\log(\varepsilon^{-1}))(1+\log(d))\}
\end{equation}
for all $\varepsilon \in (0,1]$ and $d\in\mathbb{N}$.\\
\end{definition}
Suppose $ T :(0,\infty) \times \mathbb{N} \times [0,\infty)^s \rightarrow (0,\infty)$
be defined by
\[
 T(\varepsilon^{-1},d,\vp)= \exp\{\vp(1+\log d)(1+\log\max\{\varepsilon^{-1},1\})\}
\]
Observe that this is a special case of the $T$-tractability function, where the function is polynomial in exponent $1+\log d$ and $1+\log\max\{\varepsilon^{-1},1\}$.

@@

\[
\sup_{d\in\mathbb{N}}\sum_{i=Lpd^{q_1}} \left(\min\{\lambda_{i,d},1\} \right)^{p(1+\ln d)}d^{-q} < \infty
\]
\end{example}
}

@@

>>>>

For this case where arbitrary linear functionals are allowed, the optimal solution operator is
\[
\APP_d(f,\varepsilon) = \sum_{i=1}^n \SOL(u_i) \ip[\cf_d]{f}{u_i},
\]
and the information complexity of our linear problem is
\begin{equation}\label{eq:comp_def}
\comp(\varepsilon, d) = \min \{n \in \natzero : \lambda_{n+1, d} \le \varepsilon\}, \qquad \comp:(0,\infty) \times \naturals \to \natzero.
\end{equation}

@@

Certain common notions of tractability are covered by the general situation described above.
{\small
\begin{equation*}
	\begin{array}{r@{\qquad}c@{\qquad}cc}
		\text{Tractability type} & T(\varepsilon^{-1},d,\vp)
		& T(0,d,\vp)\\
		\toprule
		\text{strong polynomial algebraic} & \max(1,\varepsilon^{-p}) & 1\\
		\text{polynomial algebraic} & \max(1,\varepsilon^{-p})d^{q} & d^{q}\\
		\text{strong polynomial exponential} &  [\max(1,\log(1 + \varepsilon^{-1})]^p & 1\\
		\text{polynomial exponential } &
		[\max(1,\log(1 + \varepsilon^{-1}))]^p  d^{q} & d^{q} \\
        \text{quasi-polynomial algebraic} & 
        \exp\{p(1+\log(\max(1,\varepsilon^{-1}))(1+\log(d))\} &
        \exp\{p(1+\log(d))\}\\
        \text{quasi-polynomial exponential} & 
        \exp\{p(1+\log(1+\log(\max(1,\varepsilon^{-1})))(1+\log(d))\} &
        \exp\{p(1+\log(d))\}
	\end{array}
\end{equation*}}

Here above, $\vp = (p,q)$ in some cases and $\vp$ is a scalar in other cases. A general reference
for results on algebraic tractability is \cite{NW08}, for exponential tractability we refer to \cite{HKW19} and \cite{KW19}.

@@

END OF THE ALTERNATIVE PROOF.

TODO: Explain that $n\ge \widetilde{J}$ is no restriction due to the upper bound on the complexity.
]

@@

\fred{Compare the lower summation indices for Theorems 1 and 2}
\begin{proof}
    \textbf{Sufficient Condition:}\\
Suppose that \eqref{eq:tractiff4} holds for some $\vp \ge 0$. By the argument in the proof of the sufficient condition for Theorem \ref{thm_main_strong_tract2}, it follows from \eqref{eq:compUP} with $d'=d$ that the information complexity has an upper bound of
\begin{equation*}
	 \comp(\varepsilon,d) \le  T(\varepsilon^{-1},d,\vp) \sum_{i=1}^\infty \frac{1}{T(\lambda_{i, d}^{-1},d,\vp)} + 1
\end{equation*}

@@

Let $\widetilde{J}:=J(\vp,d): = \lceil L_{\vp}\, T(0,d,\vp)\rceil$. By the argument in the proof of the sufficient condition for Theorem \ref{thm_main_strong_tract2}, it follows from \eqref{eq:compUP} with $d'=d$ that the information complexity has an upper bound of
\begin{equation*}
	 \comp(\varepsilon,d) \le  T(\varepsilon^{-1},d,\vp) \sum_{i=J(\vp,d)}^\infty \frac{1}{T(\lambda_{i, d}^{-1},d,\vp)} + J(\vp,d)
\end{equation*}
 Then this upper bound on the complexity can be further bounded above by
\begin{align*}
       \comp(\varepsilon,d)
       & \le T(\varepsilon^{-1},d,\vp) \Biggl [ 
       \underbrace{\sum_{i=J(\vp,d)}^\infty \frac{1}{T(\lambda_{i, d}^{-1},d,\vp)}}_{\le S_{\vp} \text{ by \eqref{eq:tractiff4}}}
       + \frac{J(\vp,d)}{T(\varepsilon^{-1},d, \vp)} \Biggr ] \\
       & \le T(\varepsilon^{-1},d,\vp)\left[ S_{\vp} + \frac{L_{\vp} T(0,d,\vp)+1}{T(\varepsilon^{-1},d, \vp)}\right]\\
       & \le T(\varepsilon^{-1},d,\vp)\left[S_{\vp} +\frac{L_{\vp} T(0,d,\vp)}{T(\varepsilon^{-1},d, \vp)}
       + \frac{1}{T(0,1, \vp)}\right]
       \qquad \text{by \eqref{eq:Tnondecreasing}}   \\
       & \le T(\varepsilon^{-1},d,\vp) 
       \underbrace{\left [  S_{\vp} + L_{\vp} + \frac{1}{T(0,1, \vp)}\right]}_{=:C_{\vp}}
       \qquad \text{by \eqref{eq:Tnondecreasing}}  
\end{align*}

@@

Let the tractability function, $T$,
be defined by
\[
 T(\varepsilon^{-1},d,p)= \max\{1,\varepsilon^{-p}\}
 \qquad \forall \varepsilon > 0, \,  d \in \naturals, \, p \ge 0.
\]
\fred{May we pull the exponent $p$ inside the $\max$?} Then Theorem \ref{thm_main_strong_tract2} yields the  equivalent condition for strong algebraic polynomial tractability:
\[
 \sup_{d\in\naturals} \sum_{i=L_p}^\infty (1,\min\{\lambda_{i,d}^p\}) < \infty \text{ for some } p \ge 0, L_p \in \naturals.
\]
Furthermore, the optimal $p$ is the infimum of all $p$ for which the above condition holds.
This essentially recovers the result in \cite[Theorem 5.1]{NW08}.
\end{example}

\begin{example}[Algebraic polynomial tractability]
Let the tractability function, $T$,
be defined by
\[
 T(\varepsilon^{-1},d,(p,q))= (\max\{1,\varepsilon^{-p}\})\, d^q
 \qquad \forall \varepsilon > 0, \,  d \in \naturals, \, p,q \ge 0.
\]
Then Theorem \ref{thm_main_tract2} yields the  equivalent condition for algebraic polynomial tractability:
\[
 \sup_{d\in\naturals} \sum_{i=L_{p,q} d^{q}}^\infty (\min\{1,\lambda_{i,d}^p\}) d^{-q} < \infty \text{ for some } p,q \ge 0, L_{p,q} \in \naturals.
\]
Furthermore, the set of optimal $(p,q)$ is a subset of
\begin{multline*}
\{(p^*,q^*) : \text{the above condition holds for all }(p,q) \in (p^*,\infty) \times (q^*,\infty) \\
\text{but not for any }(p,q) \in [0,p^*/2) \times [0,q^*/2)\}.
\end{multline*}
This essentially recovers the result on polynomial tractability in \cite[Theorem 5.1]{NW08} \fred{same theorem as above??, what about optimality in \cite[Theorem 5.1]{NW08}}.
\end{example}

\begin{example}[Strong exponential polynomial tractability]
Let the tractability function, $T$,
be defined by
\[
 T(\varepsilon^{-1},d,p)= \left[\max\{1,\log(1+\varepsilon^{-1}\})\right]^p.
\]
Then Theorem \ref{thm_main_tract2} yields the iff-condition
\[
 \sup_{d\in\naturals} \sum_{i=L_{\vp} d^{q_1}}^\infty (\min\{\lambda_{i,d},1\})^p  < \infty.
\]
\end{example}
\begin{example}[Exponential polynomial tractability]

Let the tractability function, $T$,
be defined by
\[
 T(\varepsilon^{-1},d,(p,q))=\left[\max\{1,\log(1+\varepsilon^{-1}\})\right]^p\, d^q.
\]
Then Theorem \ref{thm_main_tract2} yields the iff-condition
\[
 \sup_{d\in\naturals} \sum_{i=L_{\vp} d^{q_1}}^\infty (\min\{\lambda_{i,d},1\})^p d^{-q} < \infty.
\]
\end{example}
\bigskip

\begin{example}[Algebraic quasi-polynomial tractability]
\begin{definition}
    As defined in \cite{GW11}, we say that a problem is \emph{ quasi-polynomially tractable (QPT)} if there exist two constants $C$, $t > 0$, such that
\begin{equation}\label{expqpt}
    n(\varepsilon,d) \leq C\exp\{t(1+\log(\varepsilon^{-1}))(1+\log(d))\}
\end{equation}
for all $\varepsilon \in (0,1]$ and $d\in\mathbb{N}$.\\
\end{definition}
Let the tractability function, $T$,
be defined by
\[
 T(\varepsilon^{-1},d,p)= \exp\{p(1+\log(\max\{1,\varepsilon^{-1}\})(1+\log d)\}
\]
Observe that this is a special case of the $T$-tractability function, where the function is polynomial of exponent $1+\log d$ and $1+\log\max\{\varepsilon^{-1},1\}$.

@@

\[
\sup_{d\in\mathbb{N}}\sum_{i=Lpd^{q_1}} \left(\min\{\lambda_{i,d},1\} \right)^{p(1+\ln d)}d^{-q} < \infty
\]
\end{example}
\kachi{
\begin{example}[Separable tractability]
Let the tractability function, $T$,
be defined by
\[
 T(\varepsilon^{-1},d,(p,q))= \phi(\varepsilon^{-1},p)\, \psi(d,q)
 \qquad \forall \varepsilon > 0, \,  d \in \naturals, \, p,q \ge 0.
\]
with non-decreasing functions $\phi : (0,\infty) \times (0,\infty) \to (0,\infty)$ and $\psi : \naturals  \times (0,\infty) \to (0,\infty)$. In addition, satisfying the following conditions as in (\ref{eq:ptauassume}) 
\begin{align*}
 (\phi(\varepsilon^{-1},p))^{\tau_1} &\le K_{p,\tau_1} \phi (\varepsilon^{-1},\tau_1 p)\\
  (\psi(d,q))^{\tau_2} &\le K_{q,\tau_2} \psi (d,\tau_2 q)\\
\end{align*}
where $\tau_1, \tau_2>0$. %As the name implies, the roles of $\varepsilon^{-1}$ and $d$ are separated. 
\end{example}
\begin{example}[Non-separable symmetric tractability]
    Let the tractability function, $T$, be defined by
\[
 T(\varepsilon^{-1},d,(p,q))= \exp\{\phi(\varepsilon^{-1},p)\, \psi(d,q)\}
 \qquad \forall \varepsilon > 0, \,  d \in \naturals, \, p,q \ge 0.
\]  with non-decreasing functions $f_1$, $f_2 : [1,\infty)\rightarrow[1,\infty)$.
\end{example}
}

@@

\section{Acknowledgments}
Thank Henryk and Erich.
